A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-23 tidyindex.qmd
GGGI section revised

Frame 12 in @fig-idx-tour presents the original index values when equal weights are applied across the four dimensions. Leading the rankings, Nordic countries (Iceland, Norway, and Finland) and New Zealand show a good progression on reaching gender parity, with over 80% of the gender gap closed. In the middle range, an index value over 0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia. The last group of countries (Pakistan, Iran, Algeria, Chad), with GGGI < 0.6, have a wider gender gap to closed and Afghanistan lags further behind with GGGI = 0.4.

To observe the effect of a minor weight change on the index, we slightly vary the politics weight (while maintain equal weights among other dimensions) to generate an animation of the movement of index values in response to the changing weights. This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "local tour" [@buja2005computational; @wickham_tourr_2011].

In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index. When the weight of politics is reduced (Frame 1 and 6 vs. Frame 12), the difference in GGGI values between the Nordic and other countries narrows, suggesting a smaller variation among countries in achieving gender parity. Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs. Frame 12), nearly all the countries experience a decrease in GGGI values. A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.

Ideally, an index should be robust against minor changes in its construction components. Unfortunately, this is not the case with GGGI, which raises concerns about the index's quality. For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.

>>>>

@fig-idx-tour illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries. Frame 12 shows the dot plot of the original index values sorted from highest to lowest. Leading the rankings are the Nordic countries (Iceland, Norway, and Finland) and New Zealand. The index values are between 0 and 1, and indicate proportional difference between men and women, with a value of 0.8 indicating women are 80% of the way to equality of these measures. There is a gap in values between these countries and the middle group (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, Slovakia), and another big drop to the next group (Pakistan, Iran, Algeria, Chad). Afghanistan lags much further behind.

To make a simple illustration of sensitivity analysis, we slightly vary the weight for politics, between 0.07 and 0.52, while maintain equal weights among other dimensions. This can be viewed as an animation to examine change in relative index values as a response to the changing weights. This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "radial tour" (see @buja2005computational, @wickham_tourr_2011 and @radial-tour for more details].

Frames 1 and 6 show linear combinations where politics contributes less than the original. It is interesting to note that the gap between the nordic countries and the middle countries dissipates, indicating that this component was one reason for the relatively higher GGGI values of these countries. Also interesting is the large drop in value for Bangladesh.

Frames 18, 24, 29 show linear combinations where politics contributes more than the original. The most notable feature is that Bangladesh retains it's high index value whereas the other middle group countries decline, indicating that the politics score is a major component for Bangladesh's index value.

<!--In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index. When the weight of politics is reduced (Frame 1 and 6 vs. Frame 12), the difference in GGGI values between the Nordic and other countries narrows, suggesting a smaller variation among countries in achieving gender parity. Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs. Frame 12), nearly all the countries experience a decrease in GGGI values. A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.-->

Ideally, an index should be robust against minor changes in its construction components. This is not the case with GGGI, where small changes to one component lead to fairly large change in the index. The modular pipeline framework for computing the index makes it easy to conduct this type of sensitivity analysis, where one or more components are perturbed and the index re-calculated. 
One aspect of the GGGI not well-described in the Global Gender Gap Report is handling of missing values that are present in the initial variables for many countries, something that is common for this type of data. This could also be made more transparent with the dimension reduction module, by specifying an imputation method or providing warnings about missing values.

<!--For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.-->

### 2023-08-23 tidyindex.tex
GGGI section revised

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, sharp corners, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Frame 12 in Figure~\ref{fig-idx-tour} presents the original index values
when equal weights are applied across the four dimensions. Leading the
rankings, Nordic countries (Iceland, Norway, and Finland) and New
Zealand show a good progression on reaching gender parity, with over
80\% of the gender gap closed. In the middle range, an index value over
0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan,
Armenia, and Slovakia. The last group of countries (Pakistan, Iran,
Algeria, Chad), with GGGI \textless{} 0.6, have a wider gender gap to
closed and Afghanistan lags further behind with GGGI = 0.4.

To observe the effect of a minor weight change on the index, we slightly
vary the politics weight (while maintain equal weights among other
dimensions) to generate an animation of the movement of index values in
response to the changing weights. This visualization technique, which
presents a sequence of data projections, is referred to as a ``tour''
and the specific kind of tour used here to move between nearby
projections is known as a ``local tour'' (Andreas Buja et al. 2005;
Wickham et al. 2011).

In Figure~\ref{fig-idx-tour}, six frames have been chosen from the local
tour to explore the role of the politics dimension to the index. When
the weight of politics is reduced (Frame 1 and 6 vs.~Frame 12), the
difference in GGGI values between the Nordic and other countries
narrows, suggesting a smaller variation among countries in achieving
gender parity. Conversely, when the weight of politics increases (Frame
18, 24, and 29 vs.~Frame 12), nearly all the countries experience a
decrease in GGGI values. A noticeable exception to this trend is
Bangladesh, where its index value moves in the same direction as the
politics weight, which leads to an almost similar index value to the
Nordics when the politics dimension is assigned a weight of 0.52 in
Frame 29.

Ideally, an index should be robust against minor changes in its
construction components. Unfortunately, this is not the case with GGGI,
which raises concerns about the index's quality. For other indexes that
use linear combination to reduce variable dimensions, the demonstrated
local tour can be applied similarly to assess the index robustness
through observing how index values respond to perturbation of the
weights.

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Figure~\ref{fig-idx-tour} illustrates doing sensitivity analysis for
GGGI, for a subset of 16 countries. Frame 12 shows the dot plot of the
original index values sorted from highest to lowest. Leading the
rankings are the Nordic countries (Iceland, Norway, and Finland) and New
Zealand. The index values are between 0 and 1, and indicate proportional
difference between men and women, with a value of 0.8 indicating women
are 80\% of the way to equality of these measures. There is a gap in
values between these countries and the middle group (Brazil, Panama,
Poland, Bangladesh, Kazakhstan, Armenia, Slovakia), and another big drop
to the next group (Pakistan, Iran, Algeria, Chad). Afghanistan lags much
further behind.

To make a simple illustration of sensitivity analysis, we slightly vary
the weight for politics, between 0.07 and 0.52, while maintain equal
weights among other dimensions. This can be viewed as an animation to
examine change in relative index values as a response to the changing
weights. This visualization technique, which presents a sequence of data
projections, is referred to as a ``tour'' and the specific kind of tour
used here to move between nearby projections is known as a ``radial
tour'' (see Andreas Buja et al. (2005), Wickham et al. (2011) and Laa et
al. (2023) for more details{]}.

Frames 1 and 6 show linear combinations where politics contributes less
than the original. It is interesting to note that the gap between the
nordic countries and the middle countries dissipates, indicating that
this component was one reason for the relatively higher GGGI values of
these countries. Also interesting is the large drop in value for
Bangladesh.

Frames 18, 24, 29 show linear combinations where politics contributes
more than the original. The most notable feature is that Bangladesh
retains it's high index value whereas the other middle group countries
decline, indicating that the politics score is a major component for
Bangladesh's index value.

Ideally, an index should be robust against minor changes in its
construction components. This is not the case with GGGI, where small
changes to one component lead to fairly large change in the index. The
modular pipeline framework for computing the index makes it easy to
conduct this type of sensitivity analysis, where one or more components
are perturbed and the index re-calculated. One aspect of the GGGI not
well-described in the Global Gender Gap Report is handling of missing
values that are present in the initial variables for many countries,
something that is common for this type of data. This could also be made
more transparent with the dimension reduction module, by specifying an
imputation method or providing warnings about missing values.

@@

\leavevmode\vadjust pre{\hypertarget{ref-radial-tour}{}}%
Laa, Ursula, Alex Aumann, Dianne Cook, and German Valencia. 2023. {``New
and Simplified Manual Controls for Projection and Slice Tours, with
Application to Exploring Classification Boundaries in High
Dimensions.''} \emph{Journal of Computational and Graphical Statistics}
0 (0): 1--8. \url{https://doi.org/10.1080/10618600.2023.2206459}.
